Arrows for Parallel Computations

### 2017-05-18 paper/content/main.tex
make it compile

\usepackage{placeins}

@@

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \inlinecode{map} with and without load balancing, \inlinecode{map-reduce}) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\inlinecode{pipe}, \inlinecode{ring}, \inlinecode{torus}). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

@@

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:

@@

Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \inlinecode{using :: a -> Strategy a -> a}. For \inlinecode{parEvalN} this means that we can just apply the list of functions \inlinecode{[a -> b]} to the list of inputs \inlinecode{[a]} by zipping them with the application operator \inlinecode{\$}. We then evaluate this lazy list \inlinecode{[b]} according to a \inlinecode{Strategy [b]} with the \inlinecode{using :: a -> Strategy a -> a} operator. We construct this strategy with \inlinecode{parList :: Strategy a -> Strategy [a]} and \inlinecode{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \inlinecode{parEvalN} have the correct evaluation order, we annotate the computation with \inlinecode{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \inlinecode{parEvalN} computations. This is particularly necessary in circular communication topologies like in the \inlinecode{torus} or \inlinecode{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \inlinecode{pseq} during testing for this paper.

@@

\subsubsection{ParMonad}
The \inlinecode{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.

Our parallel evaluation function \inlinecode{parEvalN} can be defined by zipping the list of \inlinecode{[a -> b]} with the list of inputs \inlinecode{[a]} with the application operator \inlinecode{\$} just like with Multicore Haskell. Then, we map over this not yet evaluated lazy list of results \inlinecode{[b]} with \inlinecode{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \inlinecode{[Par (IVar b)]}, which we convert to \inlinecode{Par [IVar b]} with \inlinecode{sequenceA}. We wait for the computations to finish by mapping over the \inlinecode{IVar b}'s inside the \inlinecode{Par} monad with \inlinecode{get}. This results in \inlinecode{Par [b]}. We finally execute this process with \inlinecode{runPar} to finally get \inlinecode{[b]} again.

\textbf{\textcolor{red}{explain problems with laziness here. Problems with torus}}

@@

While Eden also comes with a monad \inlinecode{PA} for parallel evaluation, it also ships with a completely functional interface that includes
%\\
%\inlinecode{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
%\\
a \inlinecode{spawnF} function that
%This 
allows us to define \inlinecode{parEvalN} directly:

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	\subsection{Arrows}
\label{sec:arrows}
\begin{figure}[h]
	\includegraphics{images/arrow}
	\caption{Schematic depiction of an arrow}
\end{figure}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation. An arrow \inlinecode{arr a b} represents  a computation that converts an input \inlinecode{a} to an output \inlinecode{b}. This is defined in the arrow typeclass:

@@

\caption{The schematic depiction of \inlinecode{Arrow} combinators}

@@

\inlinecode{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \inlinecode{return}. The \inlinecode{>>>} operator is analogous to the monadic composition  \inlinecode{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}. Lastly, the \inlinecode{first} operator  takes the input arrow from \inlinecode{b} to \inlinecode{c} and converts it into an arrow on pairs with the second argument untouched. It allows us to to save input across arrows.

@@

The most prominent instances of this interface are regular functions \inlinecode{(->)} (Fig.~\ref{fig:arrowfn}),

@@

With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \inlinecode{first}, called \inlinecode{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),

@@

the \inlinecode{***} combinator which combines \inlinecode{first} and \inlinecode{second} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})

@@

and the \inlinecode{\&\&\&} combinator that constructs an arrow which outputs two different values like \inlinecode{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).

@@

A short example given by Hughes on how to use this is \inlinecode{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.

@@

% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \inlinecode{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.

@@

% \textit{Note: In the definitions Hughes gave in his paper, the notation \inlinecode{a b c} for an arrow from \inlinecode{b} to \inlinecode{c} is used. We use the equivalent definition \inlinecode{arr a b} for an arrow from \inlinecode{a} to \inlinecode{b} instead, to make it easier to find the arrow type in type signatures.}

@@

\subsection{Parallel Haskells}
Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \inlinecode{Par} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our library can switch from one to other at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \inlinecode{Par} monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \inlinecode{Par} monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,brown2010ever,dieterle2010skeleton,delaEncina2011,Dieterle2013}.

@@

Algorithmic skeletons were introduced by \citet{Cole1989}.
Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,brown2010ever,lobachev-phd,Lobachev2012}, iteration skeletons \cite{Dieterle2013}. The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel. \citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.

@@

Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\inlinecode{->}. \citet{Hughes2005} is a tutorial on arrows. Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}

\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations. Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution. In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \inlinecode{parEvalN} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach. They basically use arrows to orchestrate several tasks in robotics. We propose a general interface for parallel programming, remaining completely in Haskell.

@@

As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \inlinecode{a -> b} in parallel, \inlinecode{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
Translating this into arrow terms gives us a new operator \inlinecode{parEvalN} that lifts a list of arrows \inlinecode{[arr a b]} to a parallel arrow \inlinecode{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \inlinecode{listApp} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).

@@

With this definition of \inlinecode{parEvalN}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \inlinecode{arr} and we want this to be an interface for different backends, we introduce a new typeclass \inlinecode{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).

@@

Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \inlinecode{conf} parameter to the function. We also do not want \inlinecode{conf} to be a fixed type, as the configuration parameters can differ for different instances of \inlinecode{ArrowParallel}. So we add it to the type signature of the typeclass as well and get \inlinecode{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).

@@

Note that we don't require the \inlinecode{conf} parameter in every implementation. If it is not needed, we usually just default the \inlinecode{conf} type parameter to () and even blank it out in the parameter list of the implemented \inlinecode{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.

@@

The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \inlinecode{withStrategy :: Strategy a -> a -> a} and \inlinecode{pseq :: a -> b -> b} combinators from Multicore Haskell, where \inlinecode{withStrategy} is the same as \inlinecode{using :: a -> Strategy a -> a} but with flipped parameters.

@@

For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \inlinecode{Strategy} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \inlinecode{Conf a} data-type that simply wraps a \inlinecode{Strategy a} (Fig.~\ref{fig:confa}). We can't directly use the \inlinecode{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.

@@

To get our configurable \inlinecode{ArrowParallel} instance, we simply unwrap the strategy and pass it to \inlinecode{parList} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).

@@

The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \inlinecode{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function. The latter forks away the computation of a value and returns an \inlinecode{IVar} containing the result in the \inlinecode{Par} monad.

>>>>

% \usepackage{placeins} %% unneeded

@@

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \ensuremath{\Varid{map}} with and without load balancing, \ensuremath{\Varid{map}\mathbin{-}\Varid{reduce}}) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\ensuremath{\Varid{pipe}}, \ensuremath{\Varid{ring}}, \ensuremath{\Varid{torus}}). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

@@

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:

@@

Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \ensuremath{\Varid{parEvalN}} with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}. For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}. % $
We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator. We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations. This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.

@@

\subsubsection{ParMonad}
The \ensuremath{\Conid{Par}} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.

Our parallel evaluation function \ensuremath{\Varid{parEvalN}} can be defined by zipping the list of \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} with the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} with the application operator \ensuremath{\mathbin{\$}} just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} to transform them to a list of not yet evaluated forked away computations \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we convert to \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{sequenceA}}. We wait for the computations to finish by mapping over the \ensuremath{\Conid{IVar}\;\Varid{b}} values inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{get}}. This results in \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. We finally execute this process with \ensuremath{\Varid{runPar}} to finally get \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} again.

\mbcomment{explain problems with laziness here. Problems with torus}

@@

While Eden also comes with a monad \ensuremath{\Conid{PA}} for parallel evaluation, it also ships with a completely functional interface that includes
%\\
a \ensuremath{\Varid{spawnF}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}
%\\
%a |spawnF| 
function that
%This 
allows us to define \ensuremath{\Varid{parEvalN}} directly:

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	\subsection{Arrows}
\label{sec:arrows}
\begin{figure}[h]
	\includegraphics{images/arrow}
	\caption{Schematic depiction of an arrow}
\end{figure}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation. An arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} represents  a computation that converts an input \ensuremath{\Varid{a}} to an output \ensuremath{\Varid{b}}. This is defined in the arrow typeclass:

@@

\caption{The schematic depiction of \ensuremath{\Conid{Arrow}} combinators}

@@

\ensuremath{\Varid{arr}} is used to lift an ordinary function to an arrow type, similarly to the monadic \ensuremath{\Varid{return}}. The \ensuremath{\mathbin{>\!\!>\!\!>}} operator is analogous to the monadic composition  \ensuremath{\bind } and combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}}. Lastly, the \ensuremath{\Varid{first}} operator  takes the input arrow from \ensuremath{\Varid{b}} to \ensuremath{\Varid{c}} and converts it into an arrow on pairs with the second argument untouched. It allows us to to save input across arrows.

@@

The most prominent instances of this interface are regular functions \ensuremath{(\to )} (Fig.~\ref{fig:arrowfn}),

@@

With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),

@@

the \ensuremath{\mathbin{\*\!\*\!\*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})

@@

and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{\*\!\*\!\*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).

@@

A short example given by Hughes on how to use this is \ensuremath{\Varid{add}} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.

@@

% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.

@@

% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used. We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}

@@

\subsection{Parallel Haskells}
Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our library can switch from one to other at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \ensuremath{\Conid{Par}} monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

@@

Algorithmic skeletons were introduced by \citet{Cole1989}.
Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}. The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel. \citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.

@@

Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} is a tutorial on arrows. Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}

\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations. Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution. In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach. They basically use arrows to orchestrate several tasks in robotics. We propose a general interface for parallel programming, remaining completely in Haskell.

@@

As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, \ensuremath{\Varid{parEvalN}} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).

@@

With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).

@@

Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \ensuremath{\Varid{conf}} parameter to the function. We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}. So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).

@@

Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation. If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to () and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the ParMonad backend.

@@

The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.

@@

For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \ensuremath{\Conid{Strategy}} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}). We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.

@@

To get our configurable \ensuremath{\Conid{ArrowParallel}} instance, we simply unwrap the strategy and pass it to \ensuremath{\Varid{parList}} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).

@@

The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function. The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.

We therefore apply each function to its corresponding input value with \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}) and then fork the computation away with \ensuremath{\Varid{arr}\;\Varid{spawnP}} inside a \ensuremath{\Varid{zipWithArr}} call. This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}. In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the ParMonad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}. The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[9]{}\hsindent{8}{}\<[17]%
\>[17]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[17]{}\hsindent{8}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend}
\label{fig:ArrowParallelParMonad}
\end{figure}
